A spatio-temporal analysis of the environmental correlates of COVID-19 incidence in Spain

### 2020-04-20 Environmental-Correlates-of-COVID19-Spain/Environmental-Correlates-of-COVID19-Spain.tex
Wrapped up the conclusions.

  \title{A spatio-temporal analysis of the environmental correlates of COVID-19
incidence in Spain}
    \author[McMaster University]{Antonio Paez\corref{1}}
   \ead{paezha@mcmaster.ca} 
    \author[Universidad Politecnica de Cartagena]{Fernando A. Lopez}
   \ead{fernando.lopez@upct.es} 
    \author[Departamento de Economia]{Tatiane Menezes}
   \ead{tatiane.menezes@ufpe.br} 
    \author[Nucleo de Pesquisa]{Renata Cavalcanti}
   \ead{renata.vcsantos@gmail.com} 
    \author[Nucleo de Pesquisa]{Maira Galdino da Rocha Pitta}
   \ead{mgrpitta@ufpe.br} 
      \address[McMaster University]{School of Geography and Earth Sciences, McMaster University, 1281 Main
St W, Hamilton, ON, L8S 4K1, Canada}
    \address[Universidad Politecnica de Cartagena]{Departamento de Metodos Cuantitativos, Ciencias Juridicas, y Lenguas
Modernas, Universidad Politecnica de Cartagena, Calle Real Numero 3,
30201, Cartagena, Murcia, Spain}
    \address[Departamento de Economia]{Departamento de Economia, Universidade Federal de Pernambuco, Av dos
Economistas, s/n - Cidade Universitária, Recife - PE, 50670-901, Brasil}
    \address[Nucleo de Pesquisa]{Núcleo de Pesquisa em Inovação Terapêutica NUPIT / UFPE, Av.
Prof.~Moraes Rego, 1235 - Cidade Universitária, Recife, PE, CEP
50670-901, Brazil}
      \cortext[1]{Corresponding Author}

@@

Please note that this paper is prepared as a reproducible research
document. The source R Markdown document, as well as all data and code
needed to reproduce/review/extend the analysis can be obtained from the
following repository:

https://github.com/paezha/covid19-environmental-correlates/tree/master/Environmental-Correlates-of-COVID19-Spain

@@

Table \ref{tab:summary-best-model} presents a summary of the results of
Model 2 (\emph{lag11}). Recall that two coefficients were constrained
and are estimated only for the first equation of the system, and thus
are assumed to be constant over time. These are the coefficients
corresponding to GDP per capita and percentage of older adults. Both
coefficients are significant at \(p<=0.05\). The sign of the coefficient
for GDP per capita is positive, which indicates that wealthier regions
tend to have a higher incidence of COVID-19. This is in line with the
idea that the epidemic started earlier in wealthier places due to their
connections to a globalized world. The sign of the coefficient for
percentage of older adults, on the other hand, is negative. As
previously discussed, the level of social contact of older adults even
under normal circumstances tends to be lower than for younger people. As
a consequence, places with larger populations of older adults appear to
have a natural level of social distancing in place. It is important to
note that this does not detract from evidence that older adults are more
vulnerable individually and in institutional settings, where their case
mortality rates are perhaps the highest of all age groups. Instead, this
result indicates that their presence in the community at large tends to
depress transmission of the virus.

Of the two other control variables, the coefficient of population
density is significant at \(p<=0.05\) in 11 equations, and not
significant in 19. The coefficient for transit is significant at
\(p<=0.10\) in 22 equations, and of those, significant at \(p<=0.05\) in
14 equations. The next four variables are environmental factors. The
coefficient for humidity is significant at \(p<=0.10\) in 20 equations,
and of those, significant at \(p<=0.05\) in 19 equations. Of the
environmental variables, temperature is the only variable that has
significant coefficients in every equation at \(p<=0.05\). Finally,
sunshine has significant coefficients at \(p<=0.10\) in 18 equations,
including 15 coefficients that are significant at \(p<=0.05\).

@@

To better understand the results, we proceed to plot the coefficients in
their temporal sequence. We begin with the evolution of of the spatial
autocorrelation coefficient (\(\rho\)) and the intercept of the model,
both of which are shown in Figure \ref{fig:results-time-1}.

@@

Figure \ref{fig:beta-controls-time} shows the temporal evolution of the
coefficients for the two control variables that were not fixed over
time, i.e., \(\log(Density)\) and \(Transit\). In Section
\ref{variables} we had anticipated a positive sign for the coefficient
of this variable. The negative sign in the results is therefore somewhat
surprising. Upon reflection, however, it seems reasonable that the sign
would be negative during the period under analysis.

@@

Figure \ref{fig:beta-environmental-time} shows the temporal evolution of
the coefficient for the three environmental variables.

@@

The effects of the different variables are summarized in Table
\ref{tab:summary-impacts-best-model}.

@@

\caption{\label{tab:summary-impacts-best-model}\label{tab:summary-impacts-best-model}Summary of direct, indirect, and total effects according to best model (lag11: lagged 11-day moving average)}
\centering
\begin{tabular}[t]{lrrrl}
\toprule
Variable & Min & Mean & Max & Note\\
\midrule
\rowcolor{gray!6}  \addlinespace[0.3em]
\multicolumn{5}{l}{\textbf{Direct Effects}}\\
\hspace{1em}log(GDPpc) & 0.681 & 0.681 & 0.681 & Constrained\\
\hspace{1em}log(Older) & -0.809 & -0.809 & -0.809 & Constrained\\
\rowcolor{gray!6}  \hspace{1em}log(Density) & -0.220 & -0.096 & 0.168 & Non-constrained\\
\hspace{1em}Transit & 0.345 & 0.517 & 0.584 & Non-constrained\\
\rowcolor{gray!6}  \hspace{1em}log(Humidity) & -1.575 & -0.545 & -0.031 & Non-constrained\\
\hspace{1em}log(Temperature) & -2.019 & -1.422 & -0.929 & Non-constrained\\
\rowcolor{gray!6}  \hspace{1em}log(Sunshine) & -0.284 & 0.095 & 0.206 & Non-constrained\\
\addlinespace[0.3em]
\multicolumn{5}{l}{\textbf{Indirect Effects}}\\
\hspace{1em}log(GDPpc) & 0.632 & 0.632 & 0.632 & Constrained\\
\rowcolor{gray!6}  \hspace{1em}log(Older) & -0.751 & -0.751 & -0.751 & Constrained\\
\hspace{1em}log(Density) & -0.032 & -0.004 & 0.156 & Non-constrained\\
\rowcolor{gray!6}  \hspace{1em}Transit & 0.008 & 0.088 & 0.338 & Non-constrained\\
\hspace{1em}log(Humidity) & -1.462 & -0.156 & -0.001 & Non-constrained\\
\rowcolor{gray!6}  \hspace{1em}log(Temperature) & -1.629 & -0.273 & -0.014 & Non-constrained\\
\hspace{1em}log(Sunshine) & -0.263 & -0.006 & 0.028 & Non-constrained\\
\rowcolor{gray!6}  \addlinespace[0.3em]
\multicolumn{5}{l}{\textbf{Total Effects}}\\
\hspace{1em}log(GDPpc) & 1.313 & 1.313 & 1.313 & Constrained\\
\hspace{1em}log(Older) & -1.560 & -1.560 & -1.560 & Constrained\\
\rowcolor{gray!6}  \hspace{1em}log(Density) & -0.244 & -0.100 & 0.324 & Non-constrained\\
\hspace{1em}Transit & 0.500 & 0.606 & 0.772 & Non-constrained\\
\rowcolor{gray!6}  \hspace{1em}log(Humidity) & -3.037 & -0.701 & -0.032 & Non-constrained\\
\hspace{1em}log(Temperature) & -3.385 & -1.695 & -0.946 & Non-constrained\\
\rowcolor{gray!6}  \hspace{1em}log(Sunshine) & -0.547 & 0.089 & 0.228 & Non-constrained\\
\bottomrule
\multicolumn{5}{l}{\textit{Note: }}\\
\multicolumn{5}{l}{Non-constrained: coefficient was allowed to vary across equations}\\
\multicolumn{5}{l}{Constrained: coefficient as constant across equations}\\
\end{tabular}
\end{table}

The temporal variation of the effects is shown in Figure
\ref{fig:plot-impacts}.

\begin{figure}
\centering
\includegraphics{Environmental-Correlates-of-COVID19-Spain_files/figure-latex/plot-impacts-1.pdf}
\caption{\label{fig:plot-impacts}Temporal evolution of effects by date;
the impact is percentage change in incidence as a result of a one
percent change in value of variable.}
\end{figure}

@@

More words go here.

Limitaciones

\begin{itemize}
\item
  La temperatura puedes ser un factor pero cabría esperar que no tuviera
  un impacto lineal. Por el contrario deberiamos esperar un `punto de
  corte': Una temperatua mantenida superior a X grados durante siete
  días sea la mejor forma de incorporarla al modelo. En nuestro modelo
  (log-log) en incremento en un 1\% de la temperatura se asocia con un
  incremento beta\% en la incidencia. Esto es lo mismo si la temperatura
  es baja que si es alta.
\item
  Idem para la humedad y horas de sol
\item
  Los datos son `provisionales'. Hay gran confusión sobre la incidencia
  real. La ausencia de test de diagnóstico PCRs al inicio de la pandemia
  (también ahora) puede desvirtuar el número de casos dianosticados.
\item
  Los datos oficiales (que tampoco son fiables) son reportados a nivel
  de Comunidades Autónomas. La recopilación de datos a nivel de
  provincial son el resultado de un esfuerzo colaborativo de
  recopilacion entre distintas fuentes (principalmente gobiernos
  locales). Nuevamente puede haber sesgos importanes.
\item
  En carácter insular de las Islas Canarias y de las Islas baleares no
  se ha tenido en cuenta.
\item
  Baleares se ha linkado artificialmente con 3/4 provincias y debería de
  haberse dejado aislada para respetar su carácter insular al definir la
  matriz W.
\item
  La incidencia depende de un estado inicial. Al inicio del estudio
  había provincias en las que la epidemia estaba muy desarrollada
  (Madrid/Alava) mientras que en otras apenas habías casos. Este hecho
  no ha sido considerado en el modelo. Decretar el confinamiento debe
  tener distintos impactos entre provincias. QUIZAR METER UNA VARAIBLE
  DUMMY CON COEF BETA CONSTANTE PARA CONTROLAR AQUELLAS PROVINCIAS CON
  MAYOR NUEMRO DE CASOS AL INICIO DEL CONFINAMIENTO.
\item
  No se ha controlado por el sistema sanitario de cada provincia. Uno de
  los principales focos de contagio han sido los hospitales y los
  centros de salud. En aquellas provincias donde se ha promovido el
  mensaje ``NO IR AL MEDICO'' han presentado menor incidencia.
\item
  ¿hasta que punto las variables de control no recojen también factores
  climáticos? por ejemplo, la gente jóven vive en el sur de España que
  ha tenido menos incidencia.
\item
  Las estaciones meteorológicas para la obtencion de datos climáticos
  has sido elegidas aleatoriamente (una para cada provincia). otra
  seleccion puede dar otros resultados. AQUI SE PODRÍA HACER EL ESFUERZO
  DE CONSIDERARLAS TODAS (1000) Y CALCULAR LA MEDIA DE LAS VARAIBLES POR
  PROVINCIA
\end{itemize}

*IDEM para la humedad. idem para sunshine

\begin{itemize}
\tightlist
\item
  HACE FALTA INCLUIR UN PLOT CON LAS CORRELACIONES DE LOS RESIDUOS PARA
  DARLE RELEVANCIA A LA ESTIMACION SUR
\end{itemize}

@@

@@

@@

@@

>>>>

  \title{A spatio-temporal analysis of the environmental correlates of COVID-19
incidence in Spain}
    \author[Some School]{Author A\corref{1}}
   \ead{author.a@example.com} 
    \author[Some Institute]{Author B}
   \ead{author.b@example.com} 
      \address[Some School]{School of the Things}
    \address[Some Institute]{Institute of Everything}
    \address[Departamento de Economia]{}
      \cortext[1]{Corresponding Author}

@@

Please note that this paper is prepared as a reproducible research
document. The source R Markdown document, as well as all data and code
needed to reproduce/review/extend the analysis can be obtained from the
following anonymous Drive folder:

https://drive.google.com/open?id=1d\_40N\_QXo2Fl14r3T3CHs84IED-B1rV\_

@@

Table \ref{tab:summary-best-model} presents a summary of the results of
Model 2 (\emph{lag11}). Recall that two coefficients were constrained
and are estimated only for the first equation of the system, and thus
are assumed to be constant over time. These are the coefficients
corresponding to GDP per capita and percentage of older adults. Both
coefficients are significant at \(p\leq0.05\). The sign of the
coefficient for GDP per capita is positive, which indicates that
wealthier regions tend to have a higher incidence of COVID-19. This is
in line with the idea that the epidemic started earlier in wealthier
places due to their connections to a globalized world. The sign of the
coefficient for percentage of older adults, on the other hand, is
negative. As previously discussed, the level of social contact of older
adults even under normal circumstances tends to be lower than for
younger people. As a consequence, places with larger populations of
older adults appear to have a natural level of social distancing in
place. It is important to note that this does not detract from evidence
that older adults are more vulnerable individually and in institutional
settings, where their case mortality rates are perhaps the highest of
all age groups. Instead, this result indicates that their presence in
the community at large tends to depress transmission of the virus.

Of the two other control variables, the coefficient of population
density is significant at \(p\leq0.05\) in 11 equations, and not
significant in 19. The coefficient for transit is significant at
\(p\leq0.10\) in 22 equations, and of those, significant at
\(p\leq0.05\) in 14 equations. The next four variables are environmental
factors. The coefficient for humidity is significant at \(p\leq0.10\) in
20 equations, and of those, significant at \(p\leq0.05\) in 19
equations. Of the environmental variables, temperature is the only
variable that has significant coefficients in every equation at
\(p\leq0.05\). Finally, sunshine has significant coefficients at
\(p\leq0.10\) in 18 equations, including 15 coefficients that are
significant at \(p\leq0.05\).

@@

To better understand the results, we proceed to plot the coefficients in
their temporal sequence. At this point it is worth recalling that the
state of emergency went into effect on March 14. In the following
figures, the periods of time indicated in yellow starting on March 14
correspond to the state of emergency, with only essential travel and
selected industrial activities allowed; the period of time in orange was
the stricter lockdown when only essential travel was allowed.

We begin our discussion with the evolution of the spatial
autocorrelation coefficient (\(\rho\)) in Figure
\ref{fig:results-time-1} (left panel). It is interesting to note that
the magnitude of the spatial autocorrelation coefficient \(\rho\)
declines over the period under analysis, and is not significant for some
days. This is an interesting result: immediately prior to the
declaration of the state of emergency, there appears to have been a
strong inter-provincial contagion effect. This situation persisted for a
few days after the lockdown, and became less strong over time. At the
end of the period under examination, the strength of this effect was
much diminished. We would expect that under full compliance with strict
lockdown conditions (meaning no inter-provincial mobility) the spatial
contagion effect would be zero. The fact that it is not is evidence of
some weak contagion effect due to restrictions to the movement between
provinces.

The intercept (right panel in Figure \ref{fig:results-time-1}) is
indicative of the variation of the incidence of COVID-19, other things
being equal. Here we see that at the incidence declines somewhat
immediately after the state of emergency, only to begin increasing again
over time. Then, the incidence declines again after the stricter
lockdown and rebounds to a lower level by April 11.

@@

Figure \ref{fig:beta-controls-time} shows the temporal evolution of the
coefficients for the two control variables that were not fixed over
time, i.e., \(\log(Density)\) (left panel) and \(Transit\) (right
panel).

In Section \ref{variables} we had anticipated a positive sign for the
coefficient of this variable, and indeed, at the beginning of the period
the coefficient is positive, albeit not significant, and then remains
non-significant for the earlier part of the lockdown. We are somewhat
surprised by the way this coefficient turns significant \emph{and}
negative in the later part of the lockdown, after April 1st. This
effect, we surmise, is the result of risk compensation, a situation
where people adapt their behavior according to the \emph{perceived}
level of risk, becoming more careful when the perceived risk is higher
and viceversa (e.g., Noland, 1995; Phillips et al., 2011; Richens et
al., 2000). Consequently, residents in high density regions may perceive
the risk of infection as being higher, and adapt their behavior
accordingly - while the opposite may be true of residents in low density
regions. Anecdotally, this seems to have been the case with some
governors in the US, who resisted stay-at-home orders citing their
sparse
\footnote{https://www.inforum.com/lifestyle/health/5025620-South-Dakota-is-not-New-York-City-Noem-defends-lack-of-statewide-COVID-19-restrictions}
and rural
populations\footnote{https://www.inforum.com/lifestyle/health/5025620-South-Dakota-is-not-New-York-City-Noem-defends-lack-of-statewide-COVID-19-restrictions}.
The coefficient for Transit is positive, as expected, but with very wide
confidence intervals, and in fact not significant in the earlier part of
the period.

@@

The evolution of the coefficients for the three environmental variables
are shown in Figure \ref{fig:beta-environmental-time}. Despite a mostly
positive simple correlation with incidence (see Figure
\ref{fig:daily-correlations}) once that we control for other factors,
humidity has a negative association with incidence of COVID-19 in Spain
(top-right panel). This is in line with the literature that describes
the viability and transmission of different viruses at higher levels of
humidity. The coefficients for temperature (top-right panel) are
consistently negative and this variable is, besides the intercept, the
only one with significant coefficients in all equations. The range of
variation of this coefficient during the period examined is
approximately between -1 and -2 - although it is important to recall
that these values should not be interpreted directly as effects; more on
this below. Finally the plot for the coefficients associated with hours
of sunshine (bottom panel) is more ambiguous: prior to the lockdown, the
coefficient was negative, but significant only at \(p\leq10\). At the
beginning of the lockdown, the coefficient is no longer significative,
before becoming \emph{positive} and significant on March 23rd. This
result stands in contrast to previous findings regarding influenza,
where more hours of sunlight reduced the strength and duration of
epidemic durations (Yu et al., 2013). A difference with previous studies
is the temporal scale of the analysis: where Yu et al.~(2013) use
monthly averages, we use daily data for a much shorter period of time.
The positive sign of sunshine may well be another instance of behavioral
adaptations, whereby compliance with lockdown orders weakens on sunny
days.

@@

The preceding discussion helps to establish the inferential
contributions of the analysis. The remaining question is, what are the
implications. As discussed in Section \ref{methods} the question of the
magnitude of the effects is not clear from the coefficients alone, since
a change to a variable in a province influences, via the contagion
effect, its neighbors. For this reason, the appropriate way to estimate
the effects is to calculate both the own and contagion effects (in other
words the direct and indirect effects, respectively), as well as their
total. A summary of the different variable effects appears in Table
\ref{tab:summary-impacts-best-model}. All effects in the table are
interpreted as percentage change in the incidence of COVID-19 as a
consequence of a one percent change in the variable. The exception is
Transit (which was not log-transformed), and this variable represents
the percentage change in incidence between provinces without and with
mass transit systems.

Two variables had temporally constrained coefficients. The estimated
effect of GDP per capita is to increase the incidence of COVID-19 by
0.675\% for each percentual increase of this variable (in €1,000s). In
our view, this is a measure of inertia, as provinces with higher GDP per
capita where among the first to see exponential growth in the pandemia.
Percentage of older adults has a negative effect, and each percentual
increase in this variable is associated with a relatively small
reduction of the incidence of less than 1\%.

The temporal variation of the effects for the rest of the variables is
shown in Figure \ref{fig:plot-impacts}. The largest positive direct
effect is Transit, and the largest direct negative effects are
temperature and humidity. The direct effect of these variables is as
follows: for each percentual point increase in temperature, there is
between a 1\% and 2\% reduction in the incidence of the disease. This
effect is amplified via contagion, as seen in the central panel in the
figure, and the indirect effect can further reduce the incidence by up
to 1.5\%. The effect of humidity is also to reduce the incidence: each
percentual point of increase in humidity is associated with a reduction
of up to 1.5\% in incidence. With the addition of the indirect effect,
the total effect of a 1\% increase in humidity is to reduce incidence by
up to 3\%. As seen in the figure, the indirect (i.e., contagion) effects
are almost as strong as the direct effects immediately before and at the
beginning of the lockdown period. However, by the end of the lockdown
period, the indirect effects have weakened considerably. The effect of
density goes from positive (+0.334) to negative (-0.244). It is crucial
to keep in mind here that this is the effect of density over a period of
\emph{lockdown} and possibly subject to risk compensation, as discussed
above. In other words, the lockdown seems to be more effective in places
with higher density. Accordingly, cities should not rush to ease
lockdowns, but rather efforts should be made to reduce or eliminate risk
compensation in less densely populated regions.

What do these effects mean? Under a situation of lockdown,
inter-regional contagion is reduced, as expected, and the effect of
temperature is estimated to reduce the incidence of COVID-19 by between
one and two percent with each percentual increase of this variable.
Without a lockdown, or a relaxed lockdown, the effect is even greater,
but only because the disease is transmitted inter-regionally! Imagine a
region that is closed to its neighbors: increases in temperature are
associated with smaller reductions in incidence - however, it does not
import or export cases. A province that is not closed and not under
lockdown will see a greater effect of temperature, but at the cost of
importing/exporting cases.

@@

\caption{\label{tab:summary-impacts-best-model}\label{tab:summary-impacts-best-model}Summary of direct, indirect, and total effects according to best model (lag11: lagged 11-day moving average)}
\centering
\begin{tabular}[t]{lrrrl}
\toprule
Variable & Min & Mean & Max & Note\\
\midrule
\rowcolor{gray!6}  \addlinespace[0.3em]
\multicolumn{5}{l}{\textbf{Direct Effects}}\\
\hspace{1em}log(GDPpc) & 0.675 & 0.675 & 0.675 & Constrained\\
\hspace{1em}log(Older) & -0.802 & -0.802 & -0.802 & Constrained\\
\rowcolor{gray!6}  \hspace{1em}log(Density) & -0.220 & -0.096 & 0.167 & Non-constrained\\
\hspace{1em}Transit & 0.342 & 0.517 & 0.584 & Non-constrained\\
\rowcolor{gray!6}  \hspace{1em}log(Humidity) & -1.562 & -0.544 & -0.031 & Non-constrained\\
\hspace{1em}log(Temperature) & -2.019 & -1.421 & -0.929 & Non-constrained\\
\rowcolor{gray!6}  \hspace{1em}log(Sunshine) & -0.281 & 0.095 & 0.206 & Non-constrained\\
\addlinespace[0.3em]
\multicolumn{5}{l}{\textbf{Indirect Effects}}\\
\hspace{1em}log(GDPpc) & 0.638 & 0.638 & 0.638 & Constrained\\
\rowcolor{gray!6}  \hspace{1em}log(Older) & -0.758 & -0.758 & -0.758 & Constrained\\
\hspace{1em}log(Density) & -0.032 & -0.004 & 0.157 & Non-constrained\\
\rowcolor{gray!6}  \hspace{1em}Transit & 0.008 & 0.089 & 0.341 & Non-constrained\\
\hspace{1em}log(Humidity) & -1.475 & -0.157 & -0.001 & Non-constrained\\
\rowcolor{gray!6}  \hspace{1em}log(Temperature) & -1.644 & -0.274 & -0.014 & Non-constrained\\
\hspace{1em}log(Sunshine) & -0.266 & -0.006 & 0.028 & Non-constrained\\
\rowcolor{gray!6}  \addlinespace[0.3em]
\multicolumn{5}{l}{\textbf{Total Effects}}\\
\hspace{1em}log(GDPpc) & 1.313 & 1.313 & 1.313 & Constrained\\
\hspace{1em}log(Older) & -1.560 & -1.560 & -1.560 & Constrained\\
\rowcolor{gray!6}  \hspace{1em}log(Density) & -0.244 & -0.100 & 0.324 & Non-constrained\\
\hspace{1em}Transit & 0.500 & 0.606 & 0.772 & Non-constrained\\
\rowcolor{gray!6}  \hspace{1em}log(Humidity) & -3.037 & -0.701 & -0.032 & Non-constrained\\
\hspace{1em}log(Temperature) & -3.385 & -1.695 & -0.946 & Non-constrained\\
\rowcolor{gray!6}  \hspace{1em}log(Sunshine) & -0.547 & 0.089 & 0.228 & Non-constrained\\
\bottomrule
\multicolumn{5}{l}{\textit{Note: }}\\
\multicolumn{5}{l}{Non-constrained: coefficient was allowed to vary across equations}\\
\multicolumn{5}{l}{Constrained: coefficient as constant across equations}\\
\end{tabular}
\end{table}

\begin{figure}
\centering
\includegraphics{Environmental-Correlates-of-COVID19-Spain_files/figure-latex/plot-impacts-1.pdf}
\caption{\label{fig:plot-impacts}Temporal evolution of direct, indirect,
and total effects by date.}
\end{figure}

@@

In this paper we have presented a spatio-temporal analysis of incidence
of COVID-19 in Spain. The analysis covers a 30-day period that begins
immediately before the state of emergency was declared in the country.
The focus of the research has been on the environmental correlates of
incidence of the disease. There is a need for more empirical evidence,
as policy makers, public health practitioners, and the public begin
planning for the months ahead at this early stage of the pandemic.

Our results offer strong support for the hypothesis that incidence of
COVID-19 at the population level is lower at higher temperatures and
levels of humidity: the estiamated effect is a reduction in the
neighborhood of 3\% percent in incidence per each 1\% increase in
temperature, and a 3\% reduction in incidence per 1\% increase in
humidity \emph{under conditions of contagion}. These reductions are
estimated to be lower when lockdown conditions are in place. The
question here seems to be whether these environmental variables can
yield a bigger reduction of \emph{more} cases, or a smaller reduction of
\emph{fewer} cases. Our control variables also offer some interesting
insights. In particular, there is evidence of behavioral adaptations in
the form of risk compensation (density) and compliance with the lockdown
(sunshine). These results offer a cautionary tale, with regards to the
effectiveness of the lockdown in more dense areas, and also the
implications for compliance with stay-at-home orders as the northern
hemisphere moves towards summer and more hours of sunshine during the
day.

A key aspect of the analysis using spatial SUR models is that we were
able to model incidence of COVID-19 as an interregional contagion
process. Here, we find that the strength of the contagion effect was
dramatically reduced by the lockdown.

Needless to say, the analysis presented here is ecological, and at the
level of population health. For this reason, the analysis does not make
any claims with respect to the effect of ultraviolet light on the virus,
but rather about transmission of the virus in the population. In this
way, the analysis does not say that the virus moves less effectively
where more people live in close proximity to each other, but rather that
humans are more contagious when they feel safe in less dense regions.
Similarly, more sunshine does not mean that the virus thrives, but
rather that humans are more contagious to each other when their behavior
adapts to this environmental condition.

It is important, in this respect, to highlight some limitations of this
study.

First, the analysis was conducted under a situation of lockdown, and
therefore, besides the results are the beginning of the period examined,
one must exercise caution when trying to extrapolate the findings to
other situations. Secondly, it is well known that there is in many
countries substantial underreporting of cases of COVID-19 due to limited
testing. However, we do not suspect systematic provincial bias in
reporting, and as long as the underreporting is consistent across units
of analysis, we do not expect biased results - even if we are aware that
the number of true cases is likely higher. Thirdly, all environmental
data are based on a single station in a province. Here, we made an
effort to select the most representative station based on the
distribution of the population. An alternative approach would be to
calculate the environmental variables based on the average of multiple
stations within each province: this would likely smooth to some extent
these variables. Finally, we defined spatial neighborhoods based on
adjacency. It would be interesting to compare other connectivity
criteria, for instance based on domestic transportation instrastructure
and services. We flag this as a matter for future research.

@@

\leavevmode\hypertarget{ref-Noland1995perceived}{}%
Noland, R.B., 1995. Perceived risk and modal choice - risk compensation
in transportation system. Accident Analysis and Prevention 27, 503--521.
doi:\href{https://doi.org/10.1016/0001-4575(94)00087-3}{10.1016/0001-4575(94)00087-3}

@@

\leavevmode\hypertarget{ref-Phillips2011risk}{}%
Phillips, R.O., Fyhri, A., Sagberg, F., 2011. Risk compensation and
bicycle helmets. Risk Analysis 31, 1187--1195.
doi:\href{https://doi.org/10.1111/j.1539-6924.2011.01589.x}{10.1111/j.1539-6924.2011.01589.x}

@@

\leavevmode\hypertarget{ref-Richens2000risk}{}%
Richens, J., Imrie, J., Copas, A., 2000. Condoms and seat belts: The
parallels and the lessons. Lancet 355, 400--403.
doi:\href{https://doi.org/10.1016/s0140-6736(99)09109-6}{10.1016/s0140-6736(99)09109-6}

@@

\leavevmode\hypertarget{ref-Yu2013characterization}{}%
Yu, H.J., Alonso, W.J., Feng, L.Z., Tan, Y., Shu, Y.L., Yang, W.Z.,
Viboud, C., 2013. Characterization of regional influenza seasonality
patterns in china and implications for vaccination strategies:
Spatio-temporal modeling of surveillance data. Plos Medicine 10, 16.
doi:\href{https://doi.org/10.1371/journal.pmed.1001552}{10.1371/journal.pmed.1001552}
